Towards Practical First-Order Model Counting

### 2024-04-25 doc/paper.tex
The rest of Section 2

\Cref{sec:logic} introduces the fundamental ideas of first-order logic and
WFOMC, and discusses the different logics used in the latter.\@ Then,
\cref{sec:algebra} introduces the terminology we use to describe the output of
the original \textsc{Crane} algorithm~\cite{DBLP:conf/kr/DilkasB23}, i.e.,
functions and equations that define them.

@@

\todo[inline]{There are three subsections now, not two.}

\subsection{First-Order Logic and WFOMC (TODO)}\label{sec:logic}

\todo[inline]{Clarify the differences between the input format and the internal
  format}

\paragraph{The \FO{} Logic (After Skolemization and in a special (prenex, CNF)
  form)}
\begin{itemize}
  \item Although existential quantifiers are supported, here we describe the
        format used internally. During preprocessing, all existential
        quantifiers are eliminated using
        Skolemization~\cite{DBLP:conf/kr/BroeckMD14}, and the formula is
        rewritten into a conjunction of clauses, each of which is in
        \emph{prenex normal form}~\cite{hinman2018fundamentals}.
  \item In the spirit of keeping sorts implicit, we always assume formulas `type
        check' with respect to sorts. For example, if $P(x)$, $P(y)$, and
        $x \ne c$ are all part of the formula (for some predicate $P$, variables
        $x$ and $y$, and constant $c$), then $x$, $y$, and $c$ all have the same
        sort.
  \item A \emph{formula} is a conjunction of clauses.
  \item A \emph{clause} is of the form $\forall x_{1} \in \Delta_{1}\text{.
        }\forall x_{2} \in \Delta_{2}\dots\text{
        }\forall x_{n} \in \Delta_{n}\text{. }\phi(x_{1}, x_{2}, \dots, x_{n})$,
        where $\phi$ is a disjunction of literals that only contain variables
        $x_{1}, \dots, x_{n}$ (and any constants).
  \item We say that a clause is a \emph{(positive) unit clause} if:
        \begin{itemize}
          \item there is only one literal with a predicate, and
          \item it is a positive literal.
        \end{itemize}
  \item A \emph{literal} is either an atom (i.e., a \emph{positive} literal) or
        its negation (i.e., a \emph{negative} literal).
  \item An \emph{atom} is either:
  \begin{itemize}
    \item $P(t_{1}, \dots, t_{m})$ for some predicate $P/m$ and terms
          $t_{1}, \dots, t_{m}$ or
    \item $x=y$ for some terms $x$ and $y$
  \end{itemize}
  \item An atom is \emph{ground} if it contains no variables (i.e., only
        constants).
  \item The \emph{arity} of a predicate is the number of arguments it takes,
        i.e., $m$ in the case of predicate $P$.
  \item When we want to denote a predicate together with its arity, we write
        $P/m$.
  \item A \emph{term} is either a variable or a constant.
  \item Throughout the paper, we use set-theoretic notation, interpreting a
        formula as a set of clauses and a clause as a set of literals. Moreover,
        for readability, clauses written on separate lines are implicitly
        conjoined.
\end{itemize}

@@

\begin{definition}[WFOMC]
  Continuing from \cref{def:model}, for every predicate $P/n$ in $\phi$, let
  $w^{+}(P), w^{-}(P) \in \mathbb{R}$ be its (positive and negative)
  \emph{weights}. Unless specified otherwise, we assume weights to be equal to
  one. The \emph{(symmetric) weighted first-order model count} (WFOMC) of $\phi$
  (with respect to $\sigma$, $w^{+}$, and $w^{-}$) is the quantity
  \[
    \sum_{M \models \phi} \prod_{P(\mathbf{t}) \in M} w^{+}(P) \prod_{\neg P(\mathbf{t}) \in M} w^{-}(P),
  \]
  where the sum is over all models of $\phi$.
\end{definition}

\begin{table*}
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{A comparison of the three logics used in WFOMC in terms of:
    \begin{enumerate*}[label=(\roman*)]
      \item the number of sorts,
      \item support for constants,
      \item the maximum number of variables,
      \item allowed quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for some predicate $P/n$ and $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k, m \in \mathbb{N}_{0}$, the latter of which can depend on the domain size,
      \item $P$ is a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

In \FO{}, each term is assigned to a \emph{sort}, and each predicate $P/n$ is
assigned to a sequence of $n$ sorts. Each sort has its corresponding domain.
Most of these assignments to sorts are typically left implicit and can be
reconstructed from the quantifiers. For example, $\forall x,y \in \Delta\text{.
}P(x, y)$ implies that variables $x$ and $y$ have the same sort. On the other
hand, $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{. } P(x, y)$
implies that $x$ and $y$ have different sorts, and it would be improper to
write, e.g., $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{.
} P(x, y) \lor x = y$. \FO{} is also the only logic to support constants,
clauses with more than two variables, and the equality predicate.

\begin{remark}
  In the case of \textsc{ForcLift} and its extensions, support for a formula as
  valid input does not imply that the algorithm will be able to compile the
  formula into a circuit or graph suitable for lifted model counting. However,
  it is known that \textsc{ForcLift} compilation is guaranteed to succeed on any
  \FO{} formula without constants and with at most two
  variables~\cite{DBLP:conf/nips/Broeck11}.
\end{remark}

Both \Ctwo{} and \UFO{} are single-sorted (i.e., variables are always quantified
over the same domain) and only include formulas with at most two variables and
no constants. Unlike \UFO{}, \Ctwo{} supports existential quantifiers, including
\emph{counting quantifiers}. For example, $\exists^{=1} x\text{. }\phi(x)$ means
that there exists \emph{exactly one} $x$ such that $\phi(x)$, and
$\exists^{\le 2} x\text{. }\phi(x)$ means that there exist \emph{at most two}
such $x$. The advantage \UFO{} brings over \Ctwo{} is the support for
\emph{(equality) cardinality constraints}. For example, $|P| = 3$ constrains all
models to have \emph{exactly three positive literals with predicate $P$}. We
further illustrate the differences among the three logics with the example
below.

\begin{example}[Functions]\label{example:functions}
  To define predicate $P$ to be an endofunction on $\Delta$, in \Ctwo{} one
  could write $\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{.
  }P(x, y)$. In \UFO{}, the same could be written as
  \begin{equation}\label[formula]{eq:functions1}
    \begin{gathered}
      \forall x, y \in \Delta\text{. }S(x) \lor \neg P(x, y)\\
      |P| = |\Delta|.
    \end{gathered}
  \end{equation}
  Although \cref{eq:functions1} has some additional models, the negative weight
  $w^{-}(S) = -1$ makes some of the counts cancel out so that the WFOMC of
  \cref{eq:functions1} is still equal to the number of endofunctions on
  $\Delta$.

  In \FO{}, we would write
  \begin{equation}\label[formula]{eq:fo}
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z,
    \end{gathered}
  \end{equation}
  first stating that each $x$ must be mapped to at least one $y$ and then adding
  that if $x$ is mapped to both $y$ and $z$, then $y=z$. Note that \cref{eq:fo}
  is written with two domains instead of one. Clearly, one can still count the
  number of endofunctions by setting $|\Gamma| = |\Delta|$. As was previously
  observed by Dilkas and Belle~\shortcite{DBLP:conf/kr/DilkasB23}, such a
  reformulation can help first-order knowledge compilation algorithms find
  efficient solutions.
\end{example}

@@

We write $\expr{}$ for an arbitrary algebraic expression. Note that some terms
have different meanings in the context of algebra as compared to logic. Here, a
\emph{constant} is a non-negative integer. Similarly, a \emph{variable} is
either a parameter of a function or a variable introduced by a summation, e.g.,
$i$ in $\sum_{i=1}^{n} \expr$. A (function) \emph{signature} is
$f(x_{1}, \dots, x_{n})$ (written $f(\mathbf{x})$ for short), where $f$ is an
$n$-ary function, and each $x_{i}$ is a variable. An \emph{equation} is
$f(\mathbf{x}) = \expr{}$, where $f(\mathbf{x})$ is a signature.

>>>>

In \cref{sec:logic}, we summarise the fundamentals of first-order logic. Then,
in \cref{sec:threelogics}, we formally define WFOMC and discuss the differences
among three variants of first-order logic used for WFOMC.\@ Finally, in
\cref{sec:algebra}, we introduce the terminology used to describe the output of
the original \textsc{Crane} algorithm, i.e., functions and equations that define
them.

@@

\subsection{First-Order Logic}\label{sec:logic}

In this section, we review the basic concepts of first-order logic as they are
used in first-order knowledge compilation algorithms. There are two key
differences between the logic used within such algorithms and the logic
supported as input. First, Skolemization~\cite{DBLP:conf/kr/BroeckMD14} is used
to eliminate existential quantifiers by introducing additional predicates (some
of which have negative weights). Second, the input formula is rewritten as a
conjunction of clauses, each of which is in \emph{prenex normal
  form}~\cite{hinman2018fundamentals}.

A \emph{formula} is a conjunction of clauses. A \emph{clause} is of the form
$\forall x_{1} \in \Delta_{1}\text{.}\forall x_{2} \in \Delta_{2}\dots\text{
}\forall x_{n} \in \Delta_{n}\text{.}\phi(x_{1}, x_{2}, \dots, x_{n})$, where
$\phi$ is a disjunction of literals that only contain variables
$x_{1}, \dots, x_{n}$ (and any constants). We say that a clause is a
\emph{(positive) unit clause} if
\begin{enumerate*}[label=(\roman*)]
  \item there is only one literal with a predicate, and
  \item it is a positive literal.
\end{enumerate*}
A \emph{literal} is either an atom (i.e., a \emph{positive} literal) or its
negation (i.e., a \emph{negative} literal). An \emph{atom} is either
\begin{enumerate*}[label=(\roman*)]
  \item $P(t_{1}, \dots, t_{m})$ for some predicate $P$ and terms
  $t_{1}, \dots, t_{m}$ (written $P(\mathbf{t})$ for short) or
  \item $x=y$ for some terms $x$ and $y$.
\end{enumerate*}
An atom is \emph{ground} if it contains no variables, i.e., only constants. The
\emph{arity} of a predicate is the number of arguments it takes, i.e., $m$ in
the case of predicate $P$ above. When we want to denote a predicate together
with its arity, we write $P/m$. A \emph{term} is either a variable or a
constant. Throughout the paper, we use set-theoretic notation, interpreting a
formula as a set of clauses and a clause as a set of literals. Moreover, for
readability, clauses written on separate lines are implicitly conjoined.

\begin{table*}
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{A comparison of the three logics used in WFOMC in terms of:
    \begin{enumerate*}[label=(\roman*)]
      \item the number of sorts,
      \item support for constants,
      \item the maximum number of variables,
      \item supported quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for some predicate $P/n$ and $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k, m \in \mathbb{N}_{0}$, the latter of which can depend on the domain size,
      \item $P$ is a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

In \FO{}, each term is assigned to a \emph{sort}, and each predicate $P/n$ is
assigned to a sequence of $n$ sorts. Each sort has its corresponding domain.
Most of these assignments to sorts are typically left implicit and can be
reconstructed from the quantifiers. For example, $\forall x,y \in \Delta\text{.
}P(x, y)$ implies that variables $x$ and $y$ have the same sort. On the other
hand, $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{. } P(x, y)$
implies that $x$ and $y$ have different sorts, and it would be improper to
write, e.g., $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{.
} P(x, y) \lor x = y$. \FO{} is also the only logic to support constants,
formulas with more than two variables, and the equality predicate.

\begin{remark}
  In the case of \textsc{ForcLift} and its extensions, support for a formula as
  valid input does not imply that the algorithm can compile the formula into a
  circuit or graph suitable for lifted model counting. However, it is known that
  \textsc{ForcLift} compilation is guaranteed to succeed on any \FO{} formula
  without constants and with at most two
  variables~\cite{DBLP:conf/nips/Broeck11}.
\end{remark}

Both \Ctwo{} and \UFO{} are single-sorted (i.e., variables are always quantified
over the same domain) and only include formulas with at most two variables and
no constants. Unlike \UFO{}, \Ctwo{} supports existential quantifiers, including
\emph{counting quantifiers}. For example, $\exists^{=1} x\text{. }\phi(x)$ means
that there exists \emph{exactly one} $x$ such that $\phi(x)$, and
$\exists^{\le 2} x\text{. }\phi(x)$ means that there exist \emph{at most two}
such $x$. The advantage \UFO{} brings over \Ctwo{} is the support for
\emph{(equality) cardinality constraints}. For example, $|P| = 3$ constrains all
models to have \emph{exactly three positive literals with predicate $P$}.

@@

\begin{definition}[WFOMC~\cite{DBLP:conf/ijcai/BroeckTMDR11}]
  Continuing from \cref{def:model}, for every predicate $P$ in $\phi$, let
  $w^{+}(P), w^{-}(P) \in \mathbb{Q}$ be its \emph{weights}. Unless specified
  otherwise, we assume weights to be equal to one. The \emph{(symmetric)
    weighted first-order model count} (WFOMC) of $\phi$ (with respect to
  $\sigma$, $w^{+}$, and $w^{-}$) is
  \[
    \sum_{M \models \phi} \prod_{P(\mathbf{t}) \in M} w^{+}(P) \prod_{\neg P(\mathbf{t}) \in M} w^{-}(P),
  \]
  where the sum is over all models of $\phi$.
\end{definition}

\begin{example}[Counting functions]\label{example:functions}
  To define predicate $P$ as an endofunction on $\Delta$, in \Ctwo{} one would
  write $\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{.
  }P(x, y)$. In \UFO{}, the same could be written as
  \begin{equation}\label[formula]{eq:functions1}
    \begin{gathered}
      \forall x, y \in \Delta\text{. }S(x) \lor \neg P(x, y)\\
      |P| = |\Delta|,
    \end{gathered}
  \end{equation}
  where $w^{-}(S) = -1$. Although \cref{eq:functions1} has more models compared
  to its counterpart in \Ctwo{}, the weight function makes it so the
  \emph{weighted} model count of \cref{eq:functions1} is still equal to the
  number of endofunctions on $\Delta$.

  Equivalently, in \FO{} we would write
  \begin{equation}\label[formula]{eq:fo}
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z,
    \end{gathered}
  \end{equation}
  first stating that each $x$ must be mapped to at least one $y$ and then adding
  that if $x$ is mapped to both $y$ and $z$, then $y=z$. Note that \cref{eq:fo}
  is written with two domains instead of one. Clearly, one can still count the
  number of endofunctions by setting $|\Gamma| = |\Delta|$. As was previously
  observed by Dilkas and Belle~\shortcite{DBLP:conf/kr/DilkasB23}, such a
  reformulation can help first-order knowledge compilation algorithms find
  efficient solutions.
\end{example}

@@

We write $\expr{}$ for an arbitrary algebraic expression. Some terms have
different meanings in the context of algebra compared to logic. Here, a
\emph{constant} is a non-negative integer. Similarly, a \emph{variable} is
either a parameter of a function or a variable introduced by a summation, e.g.,
$i$ in $\sum_{i=1}^{n} \expr$. A (function) \emph{signature} is
$f(x_{1}, \dots, x_{n})$ (written $f(\mathbf{x})$ for short), where $f$ is an
$n$-ary function, and each $x_{i}$ is a variable. An \emph{equation} is
$f(\mathbf{x}) = \expr{}$, where $f(\mathbf{x})$ is a signature.
